Privacy Analysis of a Mobile Sensor Application

### 2015-01-31 paper.tex
Rewrote IT System description.

\TODO{(later)}

@@

\TODO{RG\\
* Context: Reference model IT Security analysis\\
* Application: Travel Survey\\
* Different definitions of privacy in the literature.\\
* Study outcomes depend on privacy definition
}

\begin{LGContent}
In recent days the importance of privacy protection has been amplified
by the reports about the mass surveillance of ordinary citizens on a
global scale by the NSA and other intelligence agencies around the
world.

While aiming at the noble cause of enhancing eParticipation using
mobile technologies, Live+Gov systems do process a large variety data
that is potentially infringing the citizens privacy. The captured data
includes personal information like name, phone numbers and email
addresses and sensor data from GPS and accelerometer sensors. Also
with some applications it is possible to gather images and textual
input from the citizen.

While the collection of this data is necessary for providing the
advanced services that Live+Gov aims to deliver, at he same time, the
available raw data can be used to draw a very detailed picture of the
private life of the citizen. For instance can GPS location tracking be
used to reveal shopping habits (e.g. when a car seller is visited) and
associations to political groups (when a meeting is
attended). Accelerometer data can be used to infer medical conditions
like walking disabilities. Images can contain faces of nearby persons
to with whom the citizen is associated. All this data is highly
sensitive to the citizens privacy and can be used against the citizen
if it falls in the wrong hands.

The great importance of protecting the citizens privacy should be
apparent from these examples. The European Union, as well as many
other countries in the past, has set out a number of directives that
regulate the collection, processing and use of privacy sensitive
data. We explain the most relevant legislation in Section
\ref{sec:legal}.

The ethical aspects of privacy have been the subject of study of many
social scientists and philosophers. One scholar which is particularly
relevant in our context is Charles Fried. He investiages, why we are
intuitively so sensitive to violations of our privacy. For him privacy
is not asserted as an intrinsic value by itself, he rather sated:

\begin{quote}
Privacy is not simply an absence of information about us in the minds of others;
rather it is the control we have over information about ourselves.
\end{quote}

Fried`s study on the understanding of privacy provided a great
contribution to the research on the same term in philosophy and
computer science and despite the fact his text was published in 1970,
he already included technologies to its viewpoint (like location
monitoring) that are particularly relevant to our Context.  We explain
his theory in \ref{sec:ethics} and follow his definition of privacy in
this document.

In this document we perform a thorough analysis of how to protect the
privacy of a citizen and present several implementations that form the
core of our privacy aware Sensor Data Storage and Mining Infrastructure.

We identify six main threats for the citizens privacy, and derive
eight recommendations that should be followed in order to reduce the
associated risks for hazards.  This analysis form the guideline for
the selection of 9 measures that were implemented in our system and
documented in Chapter \ref{chap:impl}.
\end{LGContent}

@@

\subsection{A Privacy Respecting Mobile Traffic Survey}

The scenario of our privacy analysis is a mobile traffic survey that supports local travel services by collecting and evaluating mobile sensor data of their customers. The survey is executed by a trustworthy third-party service. The travel services need the personal mobile data for two reasons: firstly, they want to serve their customers individually, for example by guiding them to proximity stations, to recommend adequate travel connections, to inform about delays, or to sell electronic tickets. For this purpose, the personal data for evaluation are related to the individual customers. Secondly, the travel services want to enhance their system, in that they evaluate the demand of stations, connections, changes, ways and time spent by their customers. For this purpose, personal data are anonymized before evaluation.

\subsection{IT-Systems \& Interactions}
\label{subsubsection:it-systems}

\begin{LGContent}
This section outlines the general architecture (Figure \ref{figure:IT Systems}) of IT systems for public monitoring comparable to the Live+Gov project.
This includes a description of it technical infrastructure and the interactions between its components.

The IT infrastructure of the Live+Gov system, i.e. the Live+Gov toolkit and the customization of the software components are described in detail in various project deliverables: D4.1, D4.3, D1.1, D5.1.
In this section we give an abstraction of those systems from the perspective of WP1.

%input{./figures/IT-Systems}
\begin{figure}[h]
\centering
\includegraphics[width=0.6\textwidth]{diagrams/png/it-systems.png}

%\begin{flushleft}
%\scriptsize
%\textbf{Legend:}
%\begin{itemize}
%\itemsep1pt\parskip0pt\parsep0pt
%
%\item
%\textbf{(a) Citizen:} 
%User of the L+G client application whose privacy is at stake.
%
%\item
%\textbf{(b) Service Provider:} Provides technical infrastructure.
%
%\item
%\textbf{(c) Local Authority:} Provider of the L+G system.
%
%\item
%\textbf{(d) Mobile Device:}
%Runs the L+G client application, produces and stores data sensitive to the users privacy.
%
%\item
%\textbf{(e) Data Center:} Runs the L+G services, processes and stores user data.
%
%\item
%\textbf{(f) Report Tool:} Interface to aggregated user data.
%
%\end{itemize}
%\end{flushleft}

\caption{IT Systems}
\label{figure:IT Systems}
\end{figure}

A \emph{citizen} (a) carries a mobile device running the \emph{sensor collector} application (d).
The sensor collector application is able to collect various kinds of sensor data (accelerometer, GPS, GSM, ...) and can sent the raw data back to the \emph{data center} (c).
The sensor collector can also be able to perform certain data mining operations.

Examples of such data mining operations include human activity recognition,the detection of service lines, detection of characters and faces from images.

In the Live+Gov Project mobile devices are used in particular for the following activities:
\begin{enumerate}
\item collection of GPS samples,
\item mining of Human activities (HAR) based on accelerometer samples,
\item collection of reports consisting of an image, free text and selected categories.
\end{enumerate}

The \emph{data center} stores and processes sensor data collected with the sensor collector application. It can also take into account data obtained from third parties, like the current positions of trains.
The data center can sent mining end products (traffic jam reports, bus schedule) back to the mobile device of the citizen.

In the Live+Gov Project data centers, in particular, perform the following activities:
\begin{enumerate}
\item storage of login credentials, name, and email address for each citizen,
\item storage of collected GPS samples (user id, timestamp, GPS location),
\item storage of HAR results (user id, timestamp, HAR result),
\item detection of service lines based received GPS samples (SLD),
\item storage of SLD results (userid, timestamp, SLD result),
\item sent back SLD results to mobile device,
\item storage of received reports (user id, timestamp, report),
\item detection of inherent patterns of the received reports.
\end{enumerate}

The \emph{System Provider} (b) provides and operates technical infrastructure like the Data Center and the \emph{Reporting Tool} (f).
The Reporting Tool queries the Data Center for aggregated data to visualize in form of charts and other means suitable to help understanding of monitored citizens.

\textit{Local Authorities} (c) use the reporting tool to get information in order to understand citizen movement and improve public services.
In such systems, the most valuable information for local authorities is not the raw sensor data, but aggregated views on the mining end products.

In the Live+Gov Project the reporting tools allows, in particular the following queries:
\begin{enumerate}
\item show aggregate information about which routes citizens take starting from a given location,
\item show the average waiting time of a citizen for each bus stop,
\item show routes where citizens where running to catch a bus,
\item show locations of all reports in a given time window,
\item visualize detected patterns in the report data,
\item view individual reports.
\end{enumerate}
\end{LGContent}

@@

\subsection{Privacy and Self Data Protection}
More than one hundred years ago, 1890, the American politician Louis Brandis had specified privacy as the right to be left alone \cite{WarrenBrandeis:RightToPrivacy}. From the data processing point of view, this right is best expressed by the absence of information about a person in the mind of others. Indeed, this principle of “data minimization” is still fundamental to modern data protection legislation. The modern principles of data protection are codified by many legal systems in different countries, especially in Europe \cite{EU-directive2}, and by the US save harbor principles. However, modern data protection is more than only the absence of personal data. It is based on the personal right on self-determination over information and communication \cite{BVerfG:census}. This is, for example, very well expressed by Fried’s definition of 1970:
\begin{quote}
	Privacy is not simply an absence of information about us in the minds of others; rather it is the control we have over information about ourselves. \cite{CFried:Privacy}
\end{quote}
In order to realize this requirement in the modern IT world users are provided with user control functions, for example to decide about the access on their data for future use, and to view, modify, and delete unwanted data that had been collected. In its fundamental decision of May 13, 2014, the European Court has convicted Google to accept user demands to delete links to incorrect or irrelevant personal data in the Internet (C-131/12 – Google Spain SL). Such functions would constitute a so-called self-data-protection. It requires users to be highly aware of their rights and how to use them. Complementary, system-data-protection puts the load of data protection enforcement on the data collectors, ideally even without bothering the users. Typical system data protection functions are the deletion of personal data that are not bound to service purpose, the anonymization of personal data for research, and the abstinence from forwarding personal data to third parties. Together, system-data-protection and self-data-protection are supposed to provide a fair balance between the interests of service providers and their users.
In our research work we focus on self-data-protection according to Fried’s user control requirement \cite{CFried:Privacy}. A system data protection point of view would lead to other results, namely to a set of obligations for service providers. This is subject to further research.

@@

In Fried's definition of privacy as control over information, the specification of what constitutes such information remains open. There is a vast amount of information that relates to a person and we need to get a better understanding in order to perform a thorough analysis. To this end we use the categorization by Friedewald, Finn and Wright of 2013 called the Seven Types of Privacy \cite{7ToP}. These are an extension to the four types of privacy by Roger Clarke of 1997 \cite{RClarke:4ToP}. It is important to note, that these categories are not mutually exclusive. For instance a written email is considered personal communication as well as personal data stored on a computer. However, the categories are very helpful for a privacy analysis with a focus on self-data-protection.
The seven types of privacy are as follows:
\begin{enumerate}
	\item Privacy of the Person (with respect to body functions and body characteristics)
	\item Privacy of Behavior and Action
	\item Privacy of Communication
	\item Privacy of Data and Image
	\item Privacy of Thoughts and Feelings
	\item Privacy of Location and Space 
	\item Privacy of Association (with persons, e.g. friends, and organizations, e.g. political parties)
\end{enumerate}

@@

The goal of this chapter is to analyze and identify the threats to personal privacy that are posed by collecting, storing and processing sensor data from mobile phones. We derive concrete privacy protection measures that address the main risks involved with handling such data.

In our analysis we follow the “reference model for IT security analysis” as described in \cite{Grimm:ItSecRefModel}. It supersedes earlier efforts by e.g. \cite{Avizienis}. The reference model consists of a model and a procedure. The model organizes a common security terminology in a reasonable and practical way. The procedure describes a method for analyzing the IT system based on that model. The reference model provides four views: (1) the real world of persons and their assets, (2) the potential world of requirements and threats, (3) measures and plans specified by programs, business models and attack strategies, and, finally, (4) events of running programs, data accesses and performed attacks as well as their defense. In the following sections we apply the proposed procedures of the reference model to our scenario of a travel service that collects personal data from mobile users in order to serve the users and to enhance the service. The service intends to respect the privacy of its users.

In sections \ref{subsec:world-analysis} and \ref{subsec:pot-analysis} we apply the first two steps of the reference model \cite{Grimm:ItSecRefModel}, which are related to the views on the real world and on the measures and plans. In section \ref{subsec:priv-rec} we apply the third step of the reference model by providing specific privacy recommendations and requirements as a result of the previous analysis that the system must comply with.

@@

The first step is the \emph{world view} where all components are described in their current state. It consists of the following components: Assets, IT-Systems, Actors, Conflicts of Interests, Vulnerabilities, and Interactions.

The relevant \textbf{IT-Systems} were already described in Section \ref{subsubsection:it-systems}.
\textbf{Interactions} between assets, IT-Systems, humans, and vulnerabilites as partly described in Section \ref{subsubsection:it-systems} and will be further analyzed in Section \ref{subsec:pot-analysis}.

@@

In this scenario we focus our attention to only one asset: The privacy of the citizen. Our definition of privacy is described in detail in Section \ref{sec:privacydef},

@@

This section describes the human actors previously introduced within the IT system architecture (\ref{subsubsection:it-systems}). Although the text is only a short description, a list of the most important interests is provided for each actor.

@@

\textbf{Local Authority.}
Local authorities are public offices (ministry, agency, department, ...) or other external public entities which act as direct customers of service providers. For example a department for urban mobility, orders a system to better understand usage patterns and make improvement to the urban traffic flow. Such systems are investments, and so naturally local authorities are interested in a profitable return, like increased ticket sales.

@@

\subsubsection{Vulnerabilities}
\label{subsubsection:Vulnerabilities}
This section outlines the vulnerabilities (Figure \ref{figure:Live+Gov Vulnerabilities}) of the proposed monitoring and mining system.

@@

The \emph{Potential Analysis} displays the intended and unintended interactions of the components in the world view.

The intended interactions support the underlying business objectives. Unintended interactions can lead to threats. Obviously this provides a conflict of interest with the victim's business model, to keep the asset away from unauthorized access. From the point of view of the victim, an attack would be an unintended interaction.

The potential world view consists of the following components: Business Objectives, Threats, hances/Risk, and Security Requirements. 
\textbf{Business Objectives} of the system owner were already described in Section \ref{subsubsection:it-systems} and \ref{subsec:world-analysis}. \textbf{Security Requirement}s are included to our analysis in Section \ref{subsec:priv-rec}.

\subsubsection{Threats}

A threat is a potential interaction of the components that targets an asset.
We restrict ourselves to the case of attacks, and the asset of privacy.
An attack is an interaction that is executed by an actor in response to a conflict of interest by exploiting a vulnerability of the system. We describe threats for the citizen privacy in the system.
This list can necessarily not be complete, but we make a best effort to cover the most relevant cases.

\textbf{T1. Insufficient Control Features.}
As soon as collected data of Citizens is stored on System Provider servers, all control over that data is lost. Although, control capabilities for Citizens add to the system complexity, which could motivate to omit those. Therefore interests of Citizens and System Providers are in conflict, namely it is the Citizen's \textit{Privacy vs. System Complexity} for System Providers.

\textbf{T2. Excessive Data Mining.}
The System Provider and/or the Local Authority secretly extract more private information from the collected data, than the Citizen agreed to. But results of the mining process create no disadvantages for Citizens, because there is no disclosure to third parties. This could be the case for a System Provider, who wants to test a new product and uses the pre-existing data collection.

Thus there are two possible conflicts: \textit{Privacy vs. Financial Profit of Serivce Provider} and  \textit{Privacy vs. Buisness Intelligence of Local Authority}
The threat can be provoked by either lax data handling policies of both System Providers and Local Authorities, or a weak law enforcement of existing supervision.
But the main issue, which can lead to such threats, is again a general \textit{Missing Privacy Awareness}.

@@

\subsubsection{Chances/Risk}

>>>>

* The system has been built in Live+Gov EU project
* Analysis of Privacy aspects following a structured approach
* Derive eight privacy recommendations
* Some of them have been implemented in the service

@@

* Privacy Protection very important topic in current time Fundamental
* Fundamental trade-off between enhanced functionality and convenience
  of the service (service demands) and personal data revealed by the
  use (privacy demands).
* Challenge for service provider. Find the right balance.
* We study concrete use case of a travel-survey application.

@@

\subsubsection{IT-Systems \& Interactions}
\label{subsubsection:it-systems}

Travel surveys are regularly conducted by providers of public
transportation infrastructure in order to asses the current usage of
their services and inform future planning of e.g. bus or tram lines.
In this events a large number of citizens (e.g. 5000) is asked to keep
track of their travel patterns for a constraint period of time
(e.g. one month). Traditionally the travel patterns have been recorded
manually in the form of travel diaries. Modern smart-phones are
equipped with a wide range of sensors (in particular GPS) that allow
the recording of travel-patterns in a fully automated way.

Within the Live+Gov EU project, we have established a first
prototypical implementation of such an automated travel-survey
service. In this section we give a high level description of a generic
travel-survey service. This description along with the introduced
terminology will serve as a basis for our privacy analysis.

There are three stakeholders involved with the sysem: The
\emph{citizen}, the \emph{service provider} and the \emph{travel
  agency}. The service provider is the operator of all IT systems and
sells the travel-survey service to the travel agency.

The citizen is a volunteer who is willing to share his personal travel
patterns within the travel-survey event.  He carries a mobile
smart-phone which runs the \emph{sensor collector application}. The
application collects data from various sensors available on modern
smart-phones. In particular accelerometer- and GPS-samples are
collected. Based on the accelerometer data, the application extracts
human activities (e.g. running, walking, standing). Also, the
application can send raw-sensor data back to a central data center.

The \emph{data center} stores and processes sensor data collected with
the sensor collector application. It can also take into account data
obtained from third parties, like the current positions of trains. In
particular, the data center determines if the citizen is using public
transportation and if so which lines of service (bus, train, tram) are
used. The data center can also sent mining end products and messages
(e.g. traffic jam reports, bus schedule) back to the mobile device of
the citizen.

The \emph{service provider} provides and operates all technical
infrastructure like the data center and the sensor application. He
also operates a web-based \emph{reporting tool} that aggregates
information about travel patterns of citizens within the
travel-survey. The reporting tool is used by the travel agency to gain
insights for planning and accounting of the public transportation
infrastructure.

@@

%\TODO{MM:\\
%- Shorten\\
%- DONE - Remove Implicit Type Violations\\
%- Update figure\\
%- RG: Discuss alternative definitions of privacy\\
%}

\noindent
Defining privacy is a challenge which seems impossible. This is well put to words by Serge Gutwirth, who notes:
\begin{quote}
The notion of privacy remains out of the grasp of every academic chasing it. Even when it is cornered by such additional modifiers as `our' privacy, it still finds a way to remain elusive. \cite{Gutwirth}
\end{quote}
In this section we introduce the concept of privacy as control over information.
We specify types of information we have to deal with in the context of privacy.
And finally we outline the impact mobile sensors can have on one's privacy.

\subsection{Privacy as Control over Information}
%We may need to remove the following paragraph depinding on RG's alternative definition of privacy
%Many definitions seem to focus on the vulnerabilities of privacy.
%Thus a great amount of work is invested in finding threats to prohibit instead of describing why privacy is so valuable to us.
%This way around one could derive measures to ensure and secure its value. \cite{7ToP}

%One particular scholar, who did not follow this way of thinking is Charles Fried.
One possible definition of privacy is given by Charles Fried in his work \textit{An Anatomy of Values}\cite{CFried:Privacy}.
He questioned, why we are intuitively sensitive to privacy violations.
But he did not assert privacy as an intrinsic value by itself, he rather stated:
\begin{quote}
\textbf{Privacy is} not simply an absence of information about us in the minds of others;
rather it is \textbf{the control we have over information about ourselves}. \cite{CFried:Privacy}
\end{quote}
According to Fried privacy is our ability to create and modulate the relationships to other humans.
Which makes us intuitively sensitive for its violation, because this affects what defines our identity.
Also, the ability to build and maintain relationships is essential to society.
This is what makes privacy valuable and deserving of protection. 
\cite{sep-privacy}\cite{CFried:Privacy}

\subsubsection{Relationship Creation.}

% Privacy + Trust ~> Relationship
% ===============================
% - A shares intimate information with B
% - A trusts B not to reveal that information by not constantly monitoring B, i.e. respecting B's privacy
% - B has now the possibility to share intimate information with A

We share information of great intimacy only with persons we deem trustworthy. 
Moreover, we trust those persons to not reveal information about us to others by not constantly monitoring them, i.e. respecting their privacy.
Trust needs the possibility of unknown failure.
If we would constantly monitor our partners, they cannot fail unnoticed nor can they willingly share intimate information with us.
By trusting them, we create the possibility for them to share intimate information with us.
\cite{CFried:Privacy}

\subsubsection{Relationship Modulation.}
Fried argues, depending on the person we are interacting with, we are changing the degree of intimacy we share.
If we are to share private information with others, we are most likely aware that this action is privacy related.
In that case we are able to selectively disclose information along the two dimensions of quantity and quality.
\cite{CFried:Privacy}

\subsubsection{Alternative Definitions.}
Using Fried's anatomy of privacy for this analysis is suitable because of two points.
Despite the fact his text was published in 1970, he already included technologies to its viewpoint that did not only monitor location, but also record biometric data \cite{CFried:Privacy}. 
Secondly, the concept of control over information does not enforce data minimization.
Fried's definition complies with what we call \textit{self data protection}, where persons communicate their data self-determined.
This is the opposite of \textit{system data protection}, where the collection of certain data is prohibited.

\TODO{RG: Discuss alternative definitions of privacy}

%=======================================================================================

\begin{LGContent-Hidden}
Defining privacy is a challenge which seems impossible. This is well put to words by Serge Gutwirth, who notes:

\begin{quote}
The notion of privacy remains out of the grasp of every academic chasing it. Even when it is cornered by such additional modifiers as `our' privacy, it still finds a way to remain elusive. \cite{Gutwirth}
\end{quote}

%Many researchers seem to only ``focus on the ways in which privacy can be infringed'' \cite{7ToP}.
%Thus they try to create taxonomies of \emph{privacy harms} instead of taxonomies of \emph{privacy types}.
%Those two differ in the respect that the former focuses on threats to prohibit whereas the latter focuses on values to protect.
%So one should rather evaluate what aspects are precious about privacy and develop measures to ensure their security than only forbid single actions against it \cite{7ToP}.

Many researchers seem to only ``focus on the ways in which privacy can be infringed'' \cite{7ToP}.
Thus they invest a great amount of work in defining threats to prohibit instead of describing why privacy is so valuable to us.
This way around one could derive measures to ensure and secure its value \cite{7ToP}.
One particular scholar who does this in the context of digital monitoring is Charles Fried.
He questioned, why we are intuitively sensitive to violations of privacy.
But he did not assert privacy as an intrinsic value by itself, he rather sated:

\begin{quote}
Privacy is not simply an absence of information about us in the minds of others;
rather it is \textbf{the control we have over information about ourselves}. \cite{CFried:Privacy}
\end{quote}

According to Fried privacy is a \textit{rational context}.
Which means that one is aware of such context's existence during a rational action.
If we are to share private information with others, we are most likely aware that this action is privacy related.
In that case we are able to selectively disclose information along the two dimensions of quantity and quality.
This leads Fried to his thesis, that privacy is one's ability to create and modulate his or hers social relationships, namely: friendship, love and trust. \cite{CFried:Privacy}

Depending on the conversation partner, we change the degree of intimate information we share if it is a total stranger, colleague, close friend or lover.
With close friends or lovers we share information of great intimacy we do not share with anyone else.
Moreover, we trust those persons to not reveal information about us to others by respecting their privacy.
Trust needs the possibility of unknown failure.
If we would constantly monitor our partners, they cannot fail unnoticed nor can they willingly share that information with us.
Thus they could not trust us anymore. \cite{CFried:Privacy}

So privacy or the its possibility, according to Fried, is the foundation of our core relations: friendship, love and trust.
And thus it is valuable, because those relations are essential to human society \cite{CFried:Privacy}.
Using his anatomy of privacy as foundation of our analysis is suitable because of two points.
At first, Fried's study on the understanding of privacy provided a great contribution to the research on the same term in philosophy \cite{sep-privacy} and computer science \cite{SociotechnicalArchitectureForOnlinePrivacy}.
Secondly, despite the fact his text was published in 1970, he already included technologies to its viewpoint that did not only monitor location, but also record biometric data.
\end{LGContent-Hidden}

@@

In Fried's definition of privacy as control over information, the specification of what constitues such information remains open.
There is a vast amount of information that relates to a person and we need to get a better understanding in order to perform a thorough analysis.
To this end we use the the categorization by Friedewald, Finn and Wright \cite{7ToP} called the \textit{Seven Types of Privacy}.
The seven types of privacy are an extension to the four types of privacy by Roger Clarke \cite{RClarke:4ToP}, which are:
\textit{Privacy of the Person},
\textit{Privacy of Personal Behaviour},
\textit{Privacy of Personal Communication},
\textit{Privacy of Personal Data}.
It is important to note, that these categories do not form a taxonomy, since the categories are not mutually exclusive.
For instance a written email is considered personal communication as well as personal data stored on a computer.

Moreover, Friedewald et al. argue that Clarke's types are outdated and no longer adequate in order to describe the privacy aspect of our modern, technology driven, world.
In order to address this shortcoming they extend the former four to the now introduced seven types privacy as follows:

\begin{enumerate}

\item \textbf{Privacy of the Person}
This privacy type is generally concerned with one could best understand as biometric privacy.
Friedewald et al. paraphrase it as \emph{``\om the right to keep body functions and body characteristics \om private''}.
This includes but is not limited to measures like weight, height or shoulder width;
biometric identifiers like fingerprints and DNA sequences;
or medical conditions such as limping or having a cold.

\item \textbf{Privacy of Behaviour and Action}
This privacy type is concerned with one's activities in public as well as in private spaces.
It includes but is not limited to religious practices, political activities and sexual preferences or habits.

\item \textbf{Privacy of Communication}
This privacy type is concerned with one's communication in a broad sense.
It includes written correspondence, but also conversations conducted either vis-a-vis or via electronic devices.
Friedewald et al. put it as the right to free discussion without unknown interception by third parties.

\item \textbf{Privacy of Data and Image}
This privacy type is concerned with the secrecy of personal data, especially its automatic disclosure to other individuals and organizations.
It includes data such as paychecks, insurance information or records of public administration.
However, it also refers to pictures taken without consent and digital identifiers like IP addresses or social security numbers.

\item \textbf{Privacy of Thoughts and Feelings}
This privacy type is the counterpart to Privacy of the Person like body and mind are counterparts of one another.
Comparable to the Privacy of Data and Image, Friedewald et al. state that one's thoughts and feelings must not be automatically revealed to others.
This could simply happen by the disclosure of one's diary or by technologies which allow emotion detection through biometric means.
One's body temperature or iris reflexes might infer stress or excitation.

\item \textbf{Privacy of Location and Space}
This privacy type is concerned with one's movements in public spaces and the protection of private spaces.
Friedewald et al. qualify the first dimension as one's right to move without being identified, tracked or monitored.
The second dimension is qualified as one's general right to solitude, especially the right to the inviolability of the home.

\item \textbf{Privacy of Association}
This privacy type is also put as group privacy.
Friedewald et al. state that one must have the possibility with whomever without being recorded.
Associations like friends or organizations such as political parties must not automatically be recorded because one associates with them, and vice-versa.
\end{enumerate}

\begin{LGContent-Hidden}
In Fried's definition of privacy as control over "information about onself'', the specification of what constitues such information remains open.
There is a vast amount of information that relates to a person and we need to get a better understanding in order to perform a thorough analysis.
To this end we use the the categorization by Friedewald, Finn and Wright \cite{7ToP} called the \textit{Seven Types of Privacy}.
The seven types of privacy are an extension to the four types of privacy by Roger Clarke \cite{RClarke:4ToP}, which are:
\textit{Privacy of the Person},
\textit{Privacy of Personal Behaviour},
\textit{Privacy of Personal Communication},
\textit{Privacy of Personal Data}.
It is important to note, that this categories do not form no taxonomy, since the categories are not mutually exclusive.
For instance a written email is considered personal communication as well as personal data stored on a computer.

Moreover, Friedewald et al. argue that Clarke's taxonomy is outdated and no longer adequate in order to describe the privacy aspect of our modern, technology driven, world.
In order to address this shortcoming they extend the former four to the now introduced seven types privacy as follows:

\begin{enumerate}

\item \textbf{Privacy of the Person}
This privacy type is generally concerned with one could best understand as biometric privacy.
Friedewald et al. paraphrase it as \emph{``\om the right to keep body functions and body characteristics \om private''}.
This includes but is not limited to measures like weight, height or shoulder width;
biometric identifiers like fingerprints and DNA sequences;
or medical conditions such as limping or having a cold.

\item \textbf{Privacy of Behaviour and Action}
This privacy type is concerned with one's activities in public as well as in private spaces.
It includes but is not limited to religious practices, political activities and sexual preferences or habits.

\item \textbf{Privacy of Communication}
This privacy type is concerned with one's communication in a broad sense.
It includes written correspondence, but also conversations conducted either vis-a-vis or via electronic devices.
Friedewald et al. put it as the right to free discussion without unknown interception by third parties.

\item \textbf{Privacy of Data and Image}
This privacy type is concerned with the secrecy of personal data, especially its automatic disclosure to other individuals and organizations.
It includes data such as paychecks, insurance information or records of public administration.
However, it also refers to pictures taken without consent and digital identifiers like IP addresses or social security numbers.

\item \textbf{Privacy of Thoughts and Feelings}
This privacy type is the counterpart to Privacy of the Person like body and mind are counterparts of one another.
Comparable to the Privacy of Data and Image, Friedewald et al. state that one's thoughts and feelings must not be automatically revealed to others.
This could simply happen by the disclosure of one's diary or by technologies which allow emotion detection through biometric means.
One's body temperature or iris reflexes might infer stress or excitation.

\item \textbf{Privacy of Location and Space}
This privacy type is concerned with one's movements in public spaces and the protection of private spaces.
Friedewald et al. qualify the first dimension as one's right to move without being identified, tracked or monitored.
The second dimension is qualified as one's general right to solitude, especially the right to the inviolability of the home.

\item \textbf{Privacy of Association}
This privacy type is also put as group privacy.
Friedewald et al. state that one must have the possibility with whomever without being recorded.
Associations like friends or organizations such as political parties must not automatically be recorded because one associates with them, and vice-versa.
\end{enumerate}

@@

The goal of this chapter is to analyze and identify the threads to
personal privacy that are posed by collecting, storing and processing
sensor data from mobile phones.  We derive concrete privacy protection
measures that address the main risks involved with handling such data.

In section \ref{subsec:world-analysis} and \ref{subsec:pot-analysis} we apply the first two steps of the Reference Model for IT Security Analysis as described in \cite{Grimm:ItSecRefModel}.  It supersedes earlier efforts by
e.g. \cite{Avizienis}.  The reference model consists of a \emph{model}
and a \emph{procedure}: The model aims to organize common security terminology in a reasonable and practical way. The procedure describes a method to analysis the IT system based on that model. In these sections we give a brief overview over the reference model by applying it to our scenario.

In section \ref{subsec:priv-rec} we provide specific privacy recommendations and requirements as a result of the previous analysis that the system must comply with.

@@

The first step is the \emph{world view} where all components are described in their current state. It consists of the following components:
\begin{itemize}

	\item \textbf{Assets.} 
	Things of value to one or more stakeholders.

	\item \textbf{IT-Systems.}
	The relevant IT-Systems under study. Already described in Section \ref{subsubsection:it-systems}.

	\item \textbf{Actors.}
	All identifiable stakeholders of the system under study.

	\item \textbf{Conflicts of Interests.}
	Different actors have different interests which can be in conflict which each other.

	\item \textbf{Vulnerabilities.}
	All identifiable weaknesses in the IT-System e.g. a weak passwords.

	\item \textbf{Interactions.}
	This point captures all possible interactions between assets, IT-Systems, humans, and vulnerabilites as partly described in Section \ref{subsubsection:it-systems} and will be further analyzed in \ref{subsec:pot-analysis}.

\end{itemize}

@@

In this scenario we focus our attention to only one asset: The privacy of the citizen.
Our definition of privacy is described in detail in Section \ref{sec:privacydef} and introduced in Section \ref{sec:privacytypes} the seven different types of privacy.

@@

This section describes the human actors previously introduced within the IT system architecture (\ref{subsubsection:it-systems}).
Also the additional actor \textit{External} is introduced as a person with no special access rights. Although the text is only a short description, a list of the most important interests is provided for each actor.

@@

\textbf{Local Authority.}
Local authorities are public offices (ministry, agency, department, ...) or other external public entities which act as direct customers of service providers.
They purchase a system specialized for their needs.
For example a department for urban mobility, orders a system to better understand usage patterns and make improvement to the urban traffic flow. Such systems are investments, and so naturally local authorities are interested in a profitable return, like increased ticket sales.

@@

\subsubsection{Vulnerabilities}
\label{subsubsection:Vulnerabilities}
This section outlines the vulnerabilities (Figure \ref{figure:Live+Gov Vulnerabilities}) of the proposed monitoring and mining system.
Note that vulnerabilities are not necessarily of technical nature.
The weaknesses of IT systems are often created due to misuse or misconfiguration of the various components by one or more actors.

@@

The \emph{Potential Analysis} (or \emph{Potential View}) displays the intended and unintended interactions of the components in the world view.
The intended interactions support the underlying business objectives.
Unintended interactions lead to threats. The potential view consists of the following components:

\begin{itemize}

	\item \textbf{Business Objectives.}
	Interaction of IT-system and actors that realize a business goals of the system owner were already described in Section \ref{subsubsection:it-systems} and \ref{subsec:world-analysis}.

	\item \textbf{Threats.}
	A threat is a potential interaction that destroys or harms assets of the system.
	Concrete realizations of threats can be \emph{attacks} or \emph{accidents}.
	Attacks are executed by an actor in response to a conflict of interest.
	Accidents are harmful interactions that are not willfully caused by an actor.

	\item \textbf{Chances/Risk.}
	Evaluation of chances and risks associated to the business objectives and threats. The risk associated to a threat is its expected loss.

	\item \textbf{Security Requirements.}
	Interactions (e.g. threats) that shall not occur within the system in order to achieve its business objectives.  We included these security requirements into our analysis in Section \ref{subsec:priv-rec}.

\end{itemize}

\subsubsection{Threat Specification}

A threat is a potential interaction of the components that targets an asset.
We restrict ourselves to the case of attacks, and the asset of privacy.
An attack is an interaction that is executed by an actor in response to a conflict of interest by exploiting a vulnerability of the system.
The alternative interaction of accidents are less relevant for us, since the violation of privacy always requires an actor that takes advantage of personal data.

We describe threats for the citizen privacy in the Live+Gov system.
This list can necessarily not be complete, but we make a best effort to cover the most relevant cases.

\textbf{T1. Insufficient Control Features.}
As soon as collected data of Citizens is stored on System Provider servers, all control over that data is lost. Although, control capabilities for Citizens add to the system complexity, which could motivate to omit those. Therefore interests of Citizens and System Providers are in conflict, namely it is the Citizen's \textit{Privacy vs. System Complexity} for System Providers. This is an abstract threat provoked by a general \textit{Missing Privacy Awareness} in the minds of all actors in the Live+Gov context.

\textbf{T2. Excessive Data Mining.}
The System Provider and/or the Local Authority secretly extract more private information from the collected data, than the Citizen agreed to.
But results of the mining process create no disadvantages for Citizens, because there is no disclosure to third parties.
This could be the case for a System Provider, who wants to test a new product and uses the pre-existing data collection.
Or for a Local Authority, who wants to analyze the data collection regarding fare evasion.
However, the Citizen has not agreed to such data processing nor could he, since it is conducted secretly.
This disables a Citizen to control his data adequately.
Thus there are two possible conflicts: \textit{Privacy vs. Financial Profit of Serivce Provider} and  \textit{Privacy vs. Buisness Intelligence of Local Authority}
The threat can be provoked by either lax data handling policies of both System Providers and Local Authorities, or a weak law enforcement of existing supervision.
But the main issue, which can lead to such threats, is again a general \textit{Missing Privacy Awareness}.

@@

\subsubsection{Threat Risk Evaluation}
